---
title: "03 - geoms"
subtitle: "Data visualization in R"
author: "<br> dr. mine çetinkaya-rundel <br> duke university & rstudio"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: solarized-light
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, fig.retina = 3, dpi = 300,
  out.width = "60%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

xaringanExtra::use_panelset()
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
library(countdown)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#468189",
  secondary_color = "#9DBEBB",
  inverse_header_color = "#F4E9CD",
  black_color = "#031926",
  code_highlight_color = "#F4E9CD",
  header_font_google = google_font("Atkinson Hyperlegible"),
  text_font_google   = google_font("Atkinson Hyperlegible", "300", "300i"),
  code_font_google   = google_font("Source Code Pro"),
  code_font_size = "20px",
  title_slide_background_color = "#9DBEBB",
  title_slide_text_color = "#F4E9CD",
  base_font_size = "24px",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list("margin-block-start" = "0.4rem", 
                 "margin-block-end" = "0.4rem"),
    "h2" = list("margin-block-start" = "0.4rem", 
                 "margin-block-end" = "0.4rem"),
    "h3" = list("margin-block-start" = "0.4rem", 
                 "margin-block-end" = "0.4rem"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "85%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".hand" = list("font-family" = "'Gochi Hand', cursive",
                   "font-size" = "125%"),
    ".task" = list("padding-right"    = "10px",
                   "padding-left"     = "10px",
                   "padding-top"      = "3px",
                   "padding-bottom"   = "3px",
                   "margin-bottom"    = "6px",
                   "margin-top"       = "6px",
                   "border-left"      = "solid 5px #F1DE67",
                   "background-color" = "#F1DE6750"),
    ".pull-left" = list("width" = "49%",
                        "float" = "left"),
    ".pull-right" = list("width" = "49%",
                         "float" = "right"),
    ".pull-left-wide" = list("width" = "70%",
                             "float" = "left"),
    ".pull-right-narrow" = list("width" = "27%",
                                "float" = "right"),
    ".pull-left-narrow" = list("width" = "27%",
                               "float" = "left"),
    ".pull-right-wide" = list("width" = "70%",
                              "float" = "right") 
    )
  )
```

class: middle, inverse

# 🔗 [bit.ly/dataviz-enar-2022](https://bit.ly/dataviz-enar-2022)

To follow along with the exercises, open and make a permanent copy of the RStudio Cloud project at https://rstudio.cloud/project/3796661.

---

class: middle, inverse

# Setup

---

## Packages

```{r}
# load packages
library(tidyverse)
library(openintro)
```

---

## ggplot2 theme

```{r}
# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

---

## Figure sizing

For more on including figures in R Markdown documents with the right size, resolution, etc. the following resources are great:

- [R for Data Science - Graphics for communication](https://r4ds.had.co.nz/graphics-for-communication.html)
- [Tips and tricks for working with images and figures in R Markdown documents](https://www.zevross.com/blog/2017/06/19/tips-and-tricks-for-working-with-images-and-figures-in-r-markdown-documents/)

```{r}
# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,     # 8"
  fig.asp = 0.618,   # the golden ratio
  fig.retina = 3,    # dpi multiplier for displaying HTML output on retina
  dpi = 300,         # higher dpi, sharper image
  out.width = "60%"
)
```

---

## Data prep: new variables

From last class...

```{r}
duke_forest <- duke_forest %>%
  mutate(
    decade_built = (year_built %/% 10) * 10,
    decade_built_cat = case_when(
      decade_built <= 1940 ~ "1940 or before",
      decade_built >= 1990 ~ "1990 or after",
      TRUE                 ~ as.character(decade_built)
    ),
    decade_built_cat = factor(decade_built_cat, ordered = TRUE)
  )

duke_forest %>%
  select(year_built, decade_built, decade_built_cat)
```

---

## Data prep: summary table

From last class...

```{r}
mean_area_decade <- duke_forest %>%
  group_by(decade_built_cat) %>%
  summarise(mean_area = mean(area))

mean_area_decade
```

---

class: middle, inverse

# Aesthetic mappings in ggplot2

---

## Global vs. layer-specific aesthetics

- Aesthetic mappings can be supplied in the initial `ggplot()` call, in individual layers, or in some combination of both.

- Within each layer, you can add, override, or remove mappings.

- If you only have one layer in the plot, the way you specify aesthetics doesn’t make any difference. However, the distinction is important when you start adding additional layers.

---

## Activity: Spot the difference I

.task[
Do you expect the following plots to be the same or different? If different, how? Discuss in a pair (or group) without running the code and sketch the resulting plots based on what you think the code will produce.
]

.panelset[
.panel[.panel-name[Plots]
```{r fig.show = "hide"}
# Plot A
ggplot(duke_forest, aes(x = area, y = price, color = decade_built_cat)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5)

# Plot B
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point(aes(color = decade_built_cat), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5)
```
]
]

```{r echo = FALSE}
countdown(minutes = 5, bottom = 0)
```

---

## Activity: Spot the difference II

.task[
Do you expect the following plots to be the same or different? If different, how? Discuss in a pair (or group) without running the code and sketch the resulting plots based on what you think the code will produce.
]

.panelset[
.panel[.panel-name[Plots]
```{r fig.show = "hide"}
# Plot A
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point(aes(color = decade_built_cat))

# Plot B
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point(color = "blue")

# Plot C
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point(color = "#a493ba")
```
]
]

```{r echo = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

class: middle, inverse

# Geoms

---

## Geoms

- Geometric objects, or geoms for short, perform the actual rendering of the layer, controlling the type of plot that you create

- You can think of them as "the geometric shape used to represent the data"

---

## One variable

- Discrete:
  - `geom_bar()`: display distribution of discrete variable.

- Continuous
  - `geom_histogram()`: bin and count continuous variable, display with bars
  - `geom_density()`: smoothed density estimate
  - `geom_dotplot()`: stack individual points into a dot plot
  - `geom_freqpoly()`: bin and count continuous variable, display with lines

---

## .hand[aside...]

Always use "typewriter text" (monospace font) when writing function names, and follow with `()`, e.g., 

- `geom_freqpoly()`
- `mean()`
- `lm()`

---

## `geom_bar()`

```{r}
ggplot(duke_forest, aes(x = decade_built_cat)) +
  geom_bar()
```

---

## `geom_bar()`

```{r}
ggplot(duke_forest, aes(y = decade_built_cat)) +
  geom_bar()
```

---

## `geom_histogram()`

```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_histogram()
```

---

## `geom_histogram()` and `binwidth`

.panelset[
.panel[.panel-name[20K]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_histogram(binwidth = 20000)
```
]
.panel[.panel-name[50K]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_histogram(binwidth = 50000)
```
]
.panel[.panel-name[100K]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_histogram(binwidth = 100000)
```
]
.panel[.panel-name[200K]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_histogram(binwidth = 200000)
```
]
]

---

## `geom_density()`

```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density()
```

---

## `geom_density()` and bandwidth (`bw`)

.panelset[
.panel[.panel-name[1]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(bw = 1)
```
]
.panel[.panel-name[1000]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(bw = 1000)
```
]
.panel[.panel-name[50000]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(bw = 50000)
```
]
.panel[.panel-name[500000]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(bw = 500000)
```
]
]

---

## `geom_density()` outlines

.panelset[
.panel[.panel-name[full]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(outline.type = "full")
```
]
.panel[.panel-name[both]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(outline.type = "both")
```
]
.panel[.panel-name[upper]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(outline.type = "upper")
```
]
.panel[.panel-name[lower]
```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_density(outline.type = "lower")
```
]
]

---

## `geom_dotplot()`

.task[
What does each point represent? How are their locations determined? What do the x and y axes represent?
]

```{r fig.asp = 0.4, fig.width = 12}
ggplot(duke_forest, aes(x = price)) +
  geom_dotplot(binwidth = 50000)
```

```{r echo = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## `geom_freqpoly()`

```{r}
ggplot(duke_forest, aes(x = price)) +
  geom_freqpoly(binwidth = 50000)
```

---

## `geom_freqpoly()` for comparisons

.panelset[
.panel[.panel-name[Histogram]
```{r}
ggplot(duke_forest, aes(x = price, fill = decade_built_cat)) +
  geom_histogram(binwidth = 100000)
```
]
.panel[.panel-name[Frequency polygon]
```{r}
ggplot(duke_forest, aes(x = price, color = decade_built_cat)) +
  geom_freqpoly(binwidth = 100000, size = 1)
```
]
]

---

## Two variables - both continuous

- `geom_point()`: scatterplot
- `geom_quantile()`: smoothed quantile regression
- `geom_rug()`: marginal rug plots
- `geom_smooth()`: smoothed line of best fit
- `geom_text()`: text labels

---

## `geom_rug()`

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point() +
  geom_rug()
```

---

## `geom_rug()` on the outside

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point() +
  geom_rug(outside = TRUE) +
  coord_cartesian(clip = "off")
```

---

## `geom_rug()` on the outside, but better

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point() +
  geom_rug(outside = TRUE, sides = "tr") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(1, 1, 1, 1, "cm"))
```

---

## `geom_text()`

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_text(aes(label = bed))
```


---

## `geom_text()` and more

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_text(aes(label = bed, size = bed, color = bed))
```

---

## `geom_text()` and even more

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_text(
    aes(label = bed, size = bed, color = bed),
    show.legend = FALSE
  )
```

---

## Two variables - show distribution

- `geom_bin2d()`: bin into rectangles and count
- `geom_density2d()`: smoothed 2d density estimate
- `geom_hex()`: bin into hexagons and count

---

## `geom_hex()`

```{r}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_hex()
```

---

## `geom_hex()` and warnings

- Requires installing the [**hexbin**](https://cran.r-project.org/web/packages/hexbin/index.html) package separately!

```{r eval = FALSE}
install.packages("hexbin")
```

- Otherwise you might see

```
Warning: Computation failed in `stat_binhex()`
```

---

## Two variables

- At least one discrete
  - `geom_count()`: count number of point at distinct locations
  - `geom_jitter()`: randomly jitter overlapping points

- One continuous, one discrete
  - `geom_col()`: a bar chart of pre-computed summaries
  - `geom_boxplot()`: boxplots
  - `geom_violin()`: show density of values in each group

---

## `geom_jitter()`

.task[
How are the following three plots different?
]

.panelset[
.panel[.panel-name[Plot A]
```{r fig.asp = 0.4}
ggplot(duke_forest, aes(x = bed, y = price)) +
  geom_point()
```
]
.panel[.panel-name[Plot B]
```{r fig.asp = 0.4}
ggplot(duke_forest, aes(x = bed, y = price)) +
  geom_jitter()
```
]
.panel[.panel-name[Plot C]
```{r fig.asp = 0.4}
ggplot(duke_forest, aes(x = bed, y = price)) +
  geom_jitter()
```
]
]

```{r echo = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## `geom_jitter()` and `set.seed()`

.panelset[
.panel[.panel-name[Plot A]
```{r fig.asp = 0.5}
set.seed(1234)
ggplot(duke_forest, aes(x = bed, y = price)) +
  geom_jitter()
```
]
.panel[.panel-name[Plot B]
```{r fig.asp = 0.5}
set.seed(1234)
ggplot(duke_forest, aes(x = bed, y = price)) +
  geom_jitter()
```
]
]

---

## Two variables

- One time, one continuous
  - `geom_area()`: area plot
  - `geom_line()`: line plot
  - `geom_step()`: step plot

- Display uncertainty:
  - `geom_crossbar()`: vertical bar with center
  - `geom_errorbar()`: error bars
  - `geom_linerange()`: vertical line
  - `geom_pointrange()`: vertical line with center

- Spatial
  - `geom_map()`: fast version of `geom_polygon()` for map data (more on this later...)

---

## Average price per year built

```{r}
mean_price_year <- duke_forest %>%
  group_by(year_built) %>%
  summarise(
    n = n(),
    mean_price = mean(price),
    sd_price = sd(price)
    )

mean_price_year
```

---

## `geom_line()`

```{r}
ggplot(mean_price_year, aes(x = year_built, y = mean_price)) +
  geom_line()
```

---

## `geom_area()`

```{r}
ggplot(mean_price_year, aes(x = year_built, y = mean_price)) +
  geom_area()
```

---

## `geom_step()`

```{r}
ggplot(mean_price_year, aes(x = year_built, y = mean_price)) +
  geom_step()
```

---

## `geom_errorbar()`

.task[
Describe how this plot is constructed and what the points and the lines (error bars) correspond to.
]

.panelset[
.panel[.panel-name[Code]
```{r errorbar, fig.show = "hide"}
ggplot(mean_price_year, aes(x = year_built, y = mean_price)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_price - sd_price,
                    ymax = mean_price + sd_price))
```
]
.panel[.panel-name[Plot]
```{r ref.label = "errorbar", echo = FALSE}
```
]
]

```{r echo = FALSE}
countdown(minutes = 3, bottom = 0)
```

---

## Let's clean things up a bit!

Meet your new best friend, the [**scales**](https://scales.r-lib.org/) package!

```{r}
library(scales)
```

---

## Let's clean things up a bit!

.panelset[
```{r panelset = c(output = "Plot", source = "Code"), fig.show = "hold"}
ggplot(duke_forest, aes(x = area, y = price)) +
  geom_point(alpha = 0.6, size = 2, color = "#012169") +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  scale_y_continuous(labels = label_dollar(scale = 1/1000, suffix = "K")) +
  labs(
    x = "Area (square feet)",
    y = "Sale price (USD)",
    title = "Sale prices of homes in Duke Forest",
    subtitle = "As of November 2020",
    caption = "Source: Zillow.com"
  )
```
]
